---
title: 'Intro to Poisson Regression'
subtitle: 'Case Study: Elephants'
output:
  html_document: default
editor_options: 
  chunk_output_type: console
---

## Learning goals: 
- Describe why Poisson regression is a better choice for count data than linear regression

- Identify four conditions for Poisson regression

- Fit a Poisson regression using R

- Interpret parameters from a Poisson regression model in context

- Evaluate conditions of a Poisson regression model

> Complete 1 - 2 before class.

```{r, setup, message=FALSE}
library(tidyverse)
library(gridExtra)
library(here)

here::i_am("2-Poisson-F24/01_poisson_intro_elephants.Rmd")
elephant <- read_csv(here("2-Poisson-F24/Data/elephant.csv")) 

elephant
```

Background information about the data is given in Guided Excercise 2, from section 4.11.2 of the textbook. 

An article by Poole (1989) investigated whether mating success in male elephants increases with age and whether there is a peak age for mating success. To address this question, the research team followed 41 elephants for one year and recorded their ages and their number of matings.

# Numerical and graphical summaries of elephant age and matings:

```{r}
summary(elephant)

hist.age <- ggplot(data = elephant, aes(x = age)) + 
  geom_histogram(bins = 8)
hist.mate <- ggplot(data = elephant, aes(x = matings)) + 
  geom_histogram(bins = 8)
grid.arrange(hist.age, hist.mate, ncol=1)

ggplot(data = elephant, aes(x = age, y = matings)) + 
  geom_jitter() + 
  geom_smooth(method="lm", se = FALSE) 
```

# Why does linear regression not work for this situation?

1. Is it appropriate to use linear regression here? Evaluate the regression assumptions here.

```{r}
reg1 <- lm(matings ~ age, data=elephant)
summary(reg1)

par(mfrow= c(2,2))
plot(reg1)
```


2. A log transformation can sometimes be helpful for a right-skewed response variable. Let's investigate using the steps below. 

a. What might be problematic about logging the number of matings?
```{r}
summary(elephant$matings)
summary(log(elephant$matings))
```


b. A close approximation that is commonly used, is to add 0.5 to each value before taking the log. Using this value as the response variable, fit the regression model, and comment on the assumptions.

```{r}
summary(elephant$matings)
summary(log(elephant$matings + 0.5))

ggplot(elephant, aes(x = log(matings + 0.5))) + 
  geom_histogram()

ggplot(data = elephant, aes(x = age, y = log(matings + 0.5))) + 
  geom_jitter() + 
  geom_smooth(method="lm", se = FALSE) 
```

```{r}
# reg2 <- lm(log(matings + 0.5) ~ ?????)
# summary(reg2)

# par(mfrow= c(2,2))
# plot(reg2)
```



> We'll start here in class!

Let's compare the linear regression model and the Poisson regression model:

### Linear Regression:
$$Y \sim N(\mu, \sigma)$$
$$\mu = \beta_0 + \beta_1 Age$$

### Poisson Regression:

$$ Y \sim Poisson(\lambda)$$

$$ log(\lambda) = \beta_0 + \beta_1 Age$$



#### Comparison of assumptions: 

3. Using your textbook, identify the corresponding assumption from Poisson regression (first one is done for you). Check out section 4.2.1.

Linear relationship  of Y and X   ------> Linear relationship of log($\lambda$) and X. 

Independence --------> 

Normally distributed residuals (implies normal distribution of Y) -------> 

Equal Variance of residuals (or of Y) ------>  


When we examine the original scatterplot of matings vs. age, we see the Poisson regression assumptions make more sense:

* Slight curvature of relationship of Y and X, indicates log relationship would work.

* Matings are a count response, making it suitable for Poisson distribution.

* We see that variability in matings incresaes with expected number of matings (e.g. as age increases)


# Fitting a Poisson Regression Model

4. Use the code example below to fit a Poisson regression model
```{r}
# pois.elephant <- glm(RESPONSE ~ EXPLANATORY, data = DATASET, family = DISTRIBUTION)

```


5. As when we fit a linear regression model with a log response, it is most relevant to interpret the coefficients on the log scale. Find the exponentiated coefficients and interpret them here.


It's a little tricky, but we can use geom_function to draw this curve:

```{r}
ggplot(elephant, aes(x = age, y = matings)) + 
  geom_jitter() + 
  geom_function(fun = function(x){exp(coef(pois.elephant)[1] + coef(pois.elephant)[2]*x)}) +
  labs(title = "Poisson regression model of matings")
```


6. Is the increase in matings with age statistically significant? Find the p-value in the summary output and give an interpretation. 



7. The test in the regression output is called a  "Wald test." A likelihood ratio test (also called a "Drop in Deviance Test") is generally more accurate.... although most often the results are pretty similar.  Conduct a LRT (DDT) by fitting a model with just an intercept called `fit0` and then compare models with the `anova` function. Be sure to specify test = "Chisq"



8. Find and interpret the 95\% confidence interval for the slope on age in model `pois.elephant`. 



9. The researchers are also interested in whether there is a peak age. Add a squared age term to the model and conduct a drop in deviance test. Give an interpretation of your result.





This plot puts the quadratic and linear models on the same graph, and we can visually see that adding the quadratic term (red dashed line) does in fact, seem to make little difference in the fit.
```{r}
pois.elephant2 <- glm(matings ~ age + I(age^2), family=poisson, data=elephant)

ggplot(elephant, aes(x = age, y = matings)) + 
  geom_jitter() + 
  geom_function(fun = function(x){exp(coef(pois.elephant)[1] + coef(pois.elephant)[2]*x)}) + 
  geom_function(fun = function(x){exp(coef(pois.elephant2)[1] + coef(pois.elephant2)[2]*x + coef(pois.elephant2)[3]*x^2)}, col = "red") +
  labs(title = "Comparison of linear (black) and quadratic (red) Poisson model of matings vs age")
```


# Evaluating the Poisson Model

Some simple EDA to check the Poisson assumptions. 

a. Plot the empirical log means against the explanatory variable. Which condition
does this help us verify? 

```{r}
elephantmeans <- elephant |> 
  group_by(age) |> 
  summarize(logmean = log(mean(matings)))

ggplot(elephantmeans, aes(x = age, y = logmean)) +
  geom_point()+
  geom_smooth(method = 'lm', se = FALSE)
```



b. Compare the mean and variance of matings by age. (Code below). Which condition does this help us verify? 

```{r}
elephant |> 
  group_by(age) |>
  summarize(mean = mean(matings), variance = var(matings), n = n())
```

The last column of the table shows the number of elephants at each age. We can see that often there is only one elephant observed at a particular age, and so we cannot find a variance.
To fix this we can use the "cut_number" function. (See ?cut_number... also, ?cut_interval).

```{r}
elephant |>
  mutate(agegroup = cut_number(age, 5)) |>
  group_by(agegroup) |>
  summarize(mean = mean(matings), variance = var(matings), n = n()) 
```

