---
title: "Overdispersion"
author: "Laura Boehm Vock"
date: "9/13/2021"
output: html_document
---

Learning goals

1. Identify when an overdispersed model is necessary using graphical evidence, residual deviance and goodness of fit test. 

2. Fit a quasipoisson model in R.

3. Identify the overdispersion paramter from quasipoisson model output. 


# Intro: More Elephants
```{r, setup, message=FALSE}
library(tidyverse)
library(gridExtra) 

elephant <- read_csv("~/Stats_316_F24/Class/Data/elephant.csv") %>% 
  rename(age=AGE, matings=MATINGS)

elephant
```

We have previously examined the `elephant` dataset. An article by Poole (1989) investigated whether mating success in male elephants increases with age and whether there is a peak age for mating success. To address this question, the research team followed 41 elephants for one year and recorded their ages and their number of matings.

An examination of the mean vs variance demonstrated that the two were not exactly equal:

```{r}
elephant %>%
  mutate(agegroup = cut_number(age, 5)) %>%
  group_by(agegroup) %>%
  summarize(mean = mean(matings), variance = var(matings), n = n(), ratio = variance/mean) 

elephant %>%
  mutate(agegroup = cut_number(age, 5)) %>%
  group_by(agegroup) %>%
  summarize(mean = mean(matings), variance = var(matings)) %>%
  ggplot(aes(x = mean, y = variance)) + 
  geom_point() + 
  geom_abline(slope = 1, intercept = 0) + 
  geom_smooth(method = "lm", se = FALSE)
```


1. As you might try different binwidths to examine a distribution via a histogram, we should probably remake the above table trying different bins. Experiment with changes to cut_number or use cut_interval or cut_width instead. What patterns do you see?

```{r}
elephant %>%
  mutate(agegroup = cut_interval(age, 5)) %>%
  group_by(agegroup) %>%
  summarize(mean = mean(matings), variance = var(matings), n = n(), ratio = variance/mean) 
```

```{r}
elephant %>%
  mutate(agegroup = cut_width(age, width = 4, center = 31)) %>%
  group_by(agegroup) %>%
  summarize(mean = mean(matings), variance = var(matings), n = n(), ratio = variance/mean) 
```


2. When the variance is greater than the mean, one option is to fit a Poisson regression model with an adjustment for overdispersion. In addition to examining the empirical mean and variances (as above), we can check the ratio of the residual deviance and degrees of freedom. If the Poisson distribution is truly a good fit, these numbers will be equal. Find these values in the R output below. What do they indicate?


```{r}
pois.elephant <- glm(matings ~ age, data = elephant, family = poisson)
summary(pois.elephant)
```


3. Copy and modify the code above to create a model `quasi.elephant` using `family = quasipoisson`. This fits a model with an overdispersion adjustment. 

```{r}

```

Identify how each of the following have (or have not) changed in comparing the Poisson and Quasi-Poisson models.

a. Regression coefficient estimates.



b. Regression coefficient standard errors.



c. Regression coefficient p-values.



d. Dispersion parameter



e. Residual deviance



f. AIC




Note: The Quasi-Poisson model is also known as an "Overdispersed Poisson" or "Poisson model with Overdispersion."



4. Plot the deviance residuals vs. fitted values for the Poisson regression model. Do you notice any outliers, or evidence of nonlinearity?

```{r}
elephant.pred <- data.frame(fitted = predict(pois.elephant),
                            resid = resid(pois.elephant))

ggplot(elephant.pred, aes(x = fitted, y = resid)) + 
  geom_point() +
  geom_smooth(se = FALSE, color = "red")
```


5. Conduct a goodness of fit test for the Poisson regression model. 

```{r}
# 1 - pchisq(MODEL$deviance, MODEL$df.residual)
```


6. Would you choose a quasi-Poisson or Poisson model for these elephant data?  Are there any downsides to choosing the quasi-Poisson model?



# More practice: Soccer Goals

These data come from the raw data behind the story "2019 Women’s World Cup Predictions" https://projects.fivethirtyeight.com/2019-womens-world-cup-predictions/

More details about the data can be found in the fivethirtyeight package, for dataset `wwc_2019_matches`. For your reference, I've included the code I used to restructure the data here; we will use the dataset `wwc_games` from the Class > Data folder.

Our goal is to investigate whether the Soccer Power Index (spi) is useful for predicting the number of goals a team scores in a game.

```{r}
# library(fivethirtyeight)
# data("wwc_2019_matches")
# 
# wwc_games <- wwc_2019_matches %>%
#   mutate(game = row_number()) %>%
#   select(date, game, team1, team2, score1, score2, spi1, spi2) %>%
#   pivot_longer(-c(date, game),
#                names_to= c(".value", "team_num"),
#                names_pattern = "([a-z]*)([1-2])")
#
# wwc_games
```

*Variables:*

date
team
score - number of goals scored
spi - soccer power index (higher indicates a better team)
game and team_num provide an index. Each game is represented by 2 rows.

Read in the data, then produce our usual summary information:

```{r}
 wwc_games <- read_csv("Stats 316 F21/Class/Data/wwc_games.csv")

dim(wwc_games)
str(wwc_games)
head(wwc_games)
tail(wwc_games)
summary(wwc_games)
```


1. Based on the graphs below, what seems like a reasonable initial model for these data? Do you notice anything that may be problematic?

```{r}
ggplot(wwc_games, aes(x = score)) + 
  geom_histogram()


ggplot(wwc_games, aes(x = spi, y = score)) + 
  geom_point()
  
ggplot(wwc_games, aes(x = spi, y = log(score))) + 
  geom_jitter() + 
  geom_smooth(method = "lm")
```

2. Create a graph of mean vs variance with a blue and black line, similar to the one for elephants in the example code above. What does this suggest to you about the appropriateness of a Poisson regression model?



3. Fit a regression model with spi as the explanatory variable and score as the response. Interpret a 95% confidence interval on the slope for spi. 



4. Examine the summary output for this model. Does there appear to be evidence of overdispersion (or other lack of fit)?



5. Conduct a formal test for goodness of fit.





Lack of fit can occur for many reasons! 

- outliers
- important variables are missing, or relationships are actually nonlinear
- observations are not truly independent

If you can adapt your model to account for any of these, DO THAT FIRST. Then, if all else fails, account for overdispersion with a quasi-Poisson or negative binomial model.




6. Plot the fitted values vs residuals. Do you see any outliers or evidence of non-linearity?




7. What are some suggestions for further modeling?? Try a few possibilities.


8. Give at least one reason why it may not be appropriate to assume the observations are independent in this dataset.



# Soccer goals: Negative binomial model

Another approach to accounting for overdisperson is the Negative Binomial model. The interpretation of regression coefficients is the same as in the Poisson and Quasi-Poisson models. The distributional assumption is slightly different (negative binomial rather than Poisson), and the relationship between mean and variance is thus different (more of a quadratic relationship).


```{r}
library(pscl)
# ?glm.nb

wwc.nb <- glm.nb(score ~ spi, data = wwc_games)

summary(wwc.nb)
```




