---
title: "Negative Binomial Distribution"
author: "Laura Boehm Vock"
date: "10/8/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Learning Goals:
-  Identify how shape and variance of Negative binomial distribution changes as $r$ changes.

- Recognize the negative binomial distribution as another distribution useful for modeling count data



# How does negative binomial differ from Poisson?

Consider again the medical visits data from the homework.

The Poisson distribution depends upon only one parameter, $\lambda$, and mean = variance is implicit.

The Negative binomial distribution uses two parameters, $\mu$ and $r$, where mean = $\mu$, and variance = $\mu + \mu^2/r$. Then mean and variance can be very different! 

For a specific mean value, there is only ONE Poisson distribution, but MANY possibilities for the Negative binomial!

```{r, message = FALSE, echo = FALSE}
library(AER)
data("NMES1988")
library(tidyverse)
library(gridExtra)
```

```{r}
range(NMES1988$visits)
(meanvisits <- mean(NMES1988$visits))
(varvisits <- var(NMES1988$visits))
```

```{r, echo = FALSE}
expected <- data.frame(visits = 0:89, 
                       poisson.expected = dpois(0:89, lambda=meanvisits)*nrow(NMES1988),
                       nbinom.expected1 = dnbinom(0:89, mu = meanvisits, size = 200)*nrow(NMES1988), 
                       nbinom.expected2 = dnbinom(0:89, mu = meanvisits, size = 10)*nrow(NMES1988), 
                       nbinom.expected3 = dnbinom(0:89, mu = meanvisits, size = 2)*nrow(NMES1988), 
                       nbinom.expected4 = dnbinom(0:89, mu = meanvisits, size = 0.83)*nrow(NMES1988)) 
                       


obs.hist <- ggplot(NMES1988, aes(visits)) + 
  geom_histogram(binwidth=1) +
  ggtitle("Observed Counts, \nmean = 5.8, var = 45.7") +
  ylim(0, 750)

pois.hist <- ggplot(expected, aes(x=visits, y=poisson.expected)) +
  geom_col() +
  ggtitle("Poisson Expected Counts, \nmean = 5.8, var = 5.8")+
  ylim(0, 750)

nb1.hist <- ggplot(expected, aes(x=visits, y= nbinom.expected1)) +
  geom_col() +
  ggtitle("NegBinomial, r = 200, \nmean = 5.8, var = 5.9")+
  ylim(0, 750)

nb2.hist <- ggplot(expected, aes(x=visits, y= nbinom.expected2)) +
  geom_col() +
  ggtitle("NegBinomial, r = 10, \nmean = 5.8, var = 9.1")+
  ylim(0, 750)


nb3.hist <- ggplot(expected, aes(x=visits, y= nbinom.expected3)) +
  geom_col() +
  ggtitle("NegBinomial, r = 2, \nmean = 5.8, var = 22.4")+
  ylim(0, 750)

nb4.hist <- ggplot(expected, aes(x=visits, y= nbinom.expected4)) +
  geom_col() +
  ggtitle("NegBinomial, r = 0.83, \nmean = 5.8, var = 45.9")+
  ylim(0, 750)

grid.arrange(obs.hist, pois.hist, nb1.hist, nb2.hist, nb3.hist, nb4.hist, ncol=2)  #grid.arrange is from the gridExtra package
```


The Negative binomial model with mean and variance similar to the observed medical visits mean and variance actually matches the data fairly well!




Below I highlight just the negative binomial models; its the same graphs as above.

```{r, echo = FALSE}
grid.arrange(pois.hist, 
             nb2.hist, nb3.hist, nb4.hist, ncol=1) 
```

